Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-08-18 2-body.Rmd
discussion

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking published alignments with public molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated to TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.
<!-- However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link. -->

@@

- In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem].
  Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
  If the user is not ready to make the input tree public, tree tip labels must be first standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool. *Should we say that it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?*
  This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
  Physcraper saves a summary "csv" file with results from the taxon name standardization for advantage of the user.
  If taxon names can't be mapped, their taxonomic information can't be used in comparison analysis. They will still be used in the sequence search and phylogenetic reconstruction steps.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline.
  <!-- Discussion: (although there are plans to allow storage of unpublished trees). -->

  <!--
  **EJM: Currently 2950 studies have treebase data links. It is possible that some/many of these have multiple trees in them, but that is harder to count.** -->
  <!-- there are `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website
  ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)).   -->

@@

- Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.
<!-- Alternative DNA molecular databases could be used for this purpose, either personal or public databases, such as the European Nucleotide Archive or the DNA DataBank of Japan. The code to do this with Physcraper is not yet implemented. -->
- The next step is to identify a "search taxon" to constrain the sequence search on the GenBank database within that taxonomic group.
  The search taxon can be chosen by the user from the NCBI taxonomy.
  If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the "ingroup" (Fig. \@ref(fig:framework)).
  The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
  <!-- FIGURE RECOMMENDED: Figure \@ref(fig:search)  -->
  This is known in the OpenTree as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
  To identify the MRCAT of a group of taxon names, we use the OpenTree [taxonomic tool v3](<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>) [@rees2017automated].

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local database sequence searches.
  For remote database searches, we modified the BioPython [@cock2009biopython] BLAST function from the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.
- A constrained BLAST search is performed, in which each sequence
  in the alignment is BLASTed once against all database DNA sequences belonging to the search
  taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
  along with their corresponding metadata, i.e., their GenBank accession number.
  The full sequence for each match is downloaded from NCBI into a dedicated library within the "physcraper" folder, allowing for secondary analyses to run significantly faster.
- BLAST result sequences will be discarded if they fall outside the user set min and max length cutoffs, set as proportions of the average length without gaps of sequences in the input alignment (defaults values of 80% and 120%, respectively).
  This filtering guarantees the exclusion of whole genome sequences, which create problems in multiple sequence alignment.
  The GenBank accession numbers of sequences removed due to not meeting e-value or length cutoffs are stored in output files.
- All sequences accepted up to this point are assigned an internal identifier, and are further filtered.
- New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the unified taxonomy, or they are longer than the sequence in the input alignment.
<!-- *MK mentioned it was unclear to her if this was an NCBI or OTT taxon. I think it is the unified taxonomy, hence an OTT taxon? I think it doesn't harm to be super explicit about it*
**EJM: It checks using the OTT taxonomy, but in practice OpenTree doesn't merge ncbi taxa, so it is both/either** -->
- Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- All BLAST and filtering parameters can be customized by the user.
- Reverse, complement, and reverse-complement BLAST result sequences are identified and translated using BioPython internal functions [@cock2009biopython].
- Iterative cycles of sequence similarity search can be performed, by blasting the newly found sequences until no new sequences are found. By default only one BLAST search cycle is performed in which only sequences in the input alignment are blasted.
- New sequences passing all filtering steps are added to the "csv" taxon summary file.
- A "fasta" file containing all new filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
<!-- However, it has not been added to OpenTree's synthtetic tree yet. -->
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
<!-- we know that the sequences of the Yao2020 tree have been made available on the GenBank database. Hence, we expect that updating the Gottlieb2005 tree should at least recover the same species sampled in the Yao2020 tree. -->
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from @gottlieb2005molecular. Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to perform BLAST and RAxML analyses, with bootstrap analyses taking an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The tree is 99% resolved, however, a large portion of branches are negligibly small (133 branches < 0.00001 substitution rate units), appearing as unresolved nodes (Fig. \@ref(fig:results)).
<!-- ADD percentage of nodes with bootsrap values above 95%  and below 50%?-->
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
<!-- has less new tips (134), but  -->
and contributes phylogenetic data to 134 additional taxa within *Ilex*, from which
`r 135-68` are also in updated Gottlieb2005. @yao2020phylogeny also used ITS as a marker, but their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis. The Yao2020 tree also misses 18 taxa that were added by Physcraper. This might be caused by the method they used to download existing ITS *Ilex* sequences from GenBank, which is not fully explained in the publication.
<!-- The alignments have THIS MANY columns and THIS AMOUNT of missing data. -->
<!-- RF DISTANCE INTERPRETATION -->
<!-- ADD ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE ?-->

@@

Various phylogenetic pipelines have been designed to make sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota [@sanderson2008phylota], PyPHLAWD [@smith2009mega], and SUPERSMART [@antonelli2017toward]
focus mainly on generating full trees *de novo* for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repostories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline [@izquierdo2014pumper] also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper.

As far as we know, Physcraper is the only phylogenetic pipeline that links molecular and tree data repositories.
It can also link to various biological databases, such as GBIF, the Paleobiology Database, etc, by leveraging the integrated taxonomy from OpenTree [@rees2017automated], a resource that is allowing us to reach decade-long dreams of "exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science" [@wilson2003encyclopedia].

As such, Physcraper has the potential to rapidly (in a matter of hours)
solve challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically [@webb2010biodiversity],
obtaining trees for ecophylogenetic studies [@helmus2012phylogenetic],
sistematizing molecular (and other) databases, i.e., curating taxonomic assignations [@san2010molecular],
and generating custom species trees for ecological and evolutionary downstream analyses [@stoltzfus2013phylotastic].
Data repositories hold more information than meets the eye.
Besides the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.

Usually, initial ideas about the data are changed by new analyses.
Ideally, new understanding of the data can be registered on databases,
exposing newcomers to the most up to date knowledge about the data.

<!-- Most of them focus on efficient ways to mine the data -- getting the most homologs.
Some focus on accurate ways of mining the data -- getting real and clean homologs.
Others focus on refining the alignment. -->

 <!-- *from Robert: I think you can sell the program more here. Why is it better than the other methods? You mentioned in lab meeting that its difficult to run other programs, talk about that here, talk about the speed and other advantages* -->

>>>>

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking published alignments with public molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated to TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.

@@

- In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem].
  Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
  If the user is not ready to make the input tree public, tree tip labels must be first standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool. *Should we say that it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?*
  This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
  Physcraper saves a summary "csv" file with results from the taxon name standardization for advantage of the user.
  If taxon names can't be mapped, their taxonomic information can't be used in comparison analysis. They will still be used in the sequence search and phylogenetic reconstruction steps.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline.

@@

- Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.

- The next step is to identify a "search taxon" to constrain the sequence search on the GenBank database within that taxonomic group.
  The search taxon can be chosen by the user from the NCBI taxonomy.
  If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the "ingroup" (Fig. \@ref(fig:framework)).
  The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
  <!-- FIGURE RECOMMENDED: Figure \@ref(fig:search)  -->
  This is known in the OpenTree as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
  To identify the MRCAT of a group of taxon names, we use the OpenTree [taxonomic tool v3](<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>) [@rees2017automated].

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local database sequence searches.
  For remote database searches, we modified the BioPython [@cock2009biopython] BLAST function from the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.
- A constrained BLAST search is performed, in which each sequence
  in the alignment is BLASTed once against all database DNA sequences belonging to the search
  taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
  along with their corresponding metadata, i.e., their GenBank accession number.
  The full sequence for each match is downloaded from NCBI into a dedicated library within the "physcraper" folder, allowing for secondary analyses to run significantly faster.
- BLAST result sequences will be discarded if they fall outside the user set min and max length cutoffs, set as proportions of the average length without gaps of sequences in the input alignment (defaults values of 80% and 120%, respectively).
  This filtering guarantees the exclusion of whole genome sequences, which create problems in multiple sequence alignment.
  The GenBank accession numbers of sequences removed due to not meeting e-value or length cutoffs are stored in output files.
- All sequences accepted up to this point are assigned an internal identifier, and are further filtered.
- New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the OTT taxonomy or the NCBI taxonomy, or they are longer than the sequence in the input alignment.
- Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- All BLAST and filtering parameters can be customized by the user.
- Reverse, complement, and reverse-complement BLAST result sequences are identified and translated using BioPython internal functions [@cock2009biopython].
- Iterative cycles of sequence similarity search can be performed, by blasting the newly found sequences until no new sequences are found. By default only one BLAST search cycle is performed in which only sequences in the input alignment are blasted.
- New sequences passing all filtering steps are added to the "csv" taxon summary file.
- A "fasta" file containing all new filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from @gottlieb2005molecular. Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to perform BLAST and RAxML analyses, with bootstrap analyses taking an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The tree is 99% resolved, however, a large portion of branches are negligibly small (133 branches < 0.00001 substitution rate units), appearing as unresolved nodes (Fig. \@ref(fig:results)).
<!-- ADD percentage of nodes with bootsrap values above 95%  and below 50%?-->
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
<!-- has less new tips (134), but  -->
and contributes phylogenetic data to 134 additional taxa within *Ilex*, from which
`r 135-68` are also in updated Gottlieb2005. @yao2020phylogeny also used ITS as a marker, but their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis. The Yao2020 tree also misses 18 taxa that were added by Physcraper. This might be caused by the method they used to download existing ITS *Ilex* sequences from GenBank, which is not fully explained in the publication.
<!-- The alignments have THIS MANY columns and THIS AMOUNT of missing data. -->
<!-- RF DISTANCE INTERPRETATION -->
<!-- ADD ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE ?-->

@@

Various phylogenetic pipelines have been designed to make sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota [@sanderson2008phylota], PyPHLAWD [@smith2009mega], and SUPERSMART [@antonelli2017toward]
focus mainly on generating full trees *de novo* for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repositories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline [@izquierdo2014pumper] also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper.

As far as we know, Physcraper is the only phylogenetic pipeline that links molecular and tree data repositories.
It can also link to various biological databases, such as GBIF, the Paleobiology Database, etc, by leveraging the integrated taxonomy from OpenTree [@rees2017automated], a resource that is contributing to the Encyclopedia of Life main goal of "exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science" [@wilson2003encyclopedia].

As such, Physcraper can be used to rapidly (in a matter of hours)
solve challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically [@webb2010biodiversity],
obtaining trees for ecophylogenetic studies [@helmus2012phylogenetic],
sistematizing molecular (and other) databases, i.e., curating taxonomic assignations [@san2010molecular],
and generating custom species trees for ecological and evolutionary downstream analyses [@stoltzfus2013phylotastic].

Data repositories hold more information than meets the eye.
Besides the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.
Usually, initial ideas about the data are changed by new analyses.
Ideally, new understanding of the data can be registered on databases,
exposing newcomers to the most up to date knowledge about the data.

### 2020-08-18 2.2-methods-ideas.Rmd
some edits
discussion

>>>>

figure of unified taxonomy

a table that is a snipet of the csv to show the unified taxonoy for some taxa
also, a screen shoy of the diffeent taxon ides on opentree website works!

**EJM: Currently 2950 studies have treebase data links. It is possible that some/many of these have multiple trees in them, but that is harder to count.**
there are `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website
([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)).

Alternative DNA molecular databases could be used for this purpose, either personal or public databases, such as the European Nucleotide Archive or the DNA DataBank of Japan. The code to do this with Physcraper is not yet implemented.

### 2020-08-18 2.4-discussion-ideas.Rmd
discussion

@@

>>>>

Discussion: (although there are plans to allow storage of unpublished trees).

@@

<!-- Most of them focus on efficient ways to mine the data -- getting the most homologs.
Some focus on accurate ways of mining the data -- getting real and clean homologs.
Others focus on refining the alignment. -->

 <!-- *from Robert: I think you can sell the program more here. Why is it better than the other methods? You mentioned in lab meeting that its difficult to run other programs, talk about that here, talk about the speed and other advantages* -->

### 2020-08-18 docs/submission-mee-2020-08-18.tex
discussion

\textbf{Word count}: 3769

@@

  New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the unified taxonomy, or they are longer than the sequence in the input alignment.

@@

Various phylogenetic pipelines have been designed to make sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}), PyPHLAWD (Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}), and SUPERSMART (Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017})
focus mainly on generating full trees \emph{de novo} for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies \emph{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repostories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline (Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}) also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper.

As far as we know, Physcraper is the only phylogenetic pipeline that links molecular and tree data repositories.
It can also link to various biological databases, such as GBIF, the Paleobiology Database, etc, by leveraging the integrated taxonomy from OpenTree (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), a resource that is allowing us to reach decade-long dreams of ``exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science'' (Wilson \protect\hyperlink{ref-wilson2003encyclopedia}{2003}).

As such, Physcraper has the potential to rapidly (in a matter of hours)
solve challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically (Webb \emph{et al.} \protect\hyperlink{ref-webb2010biodiversity}{2010}),
obtaining trees for ecophylogenetic studies (Helmus \& Ives \protect\hyperlink{ref-helmus2012phylogenetic}{2012}),
sistematizing molecular (and other) databases, i.e., curating taxonomic assignations (San Mauro \& Agorreta \protect\hyperlink{ref-san2010molecular}{2010}),
and generating custom species trees for ecological and evolutionary downstream analyses (Stoltzfus \emph{et al.} \protect\hyperlink{ref-stoltzfus2013phylotastic}{2013}).
Data repositories hold more information than meets the eye.
Besides the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.

Usually, initial ideas about the data are changed by new analyses.
Ideally, new understanding of the data can be registered on databases,
exposing newcomers to the most up to date knowledge about the data.

>>>>

\textbf{Word count}: 3775

@@

  New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the OTT taxonomy or the NCBI taxonomy, or they are longer than the sequence in the input alignment.

@@

Various phylogenetic pipelines have been designed to make sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}), PyPHLAWD (Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}), and SUPERSMART (Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017})
focus mainly on generating full trees \emph{de novo} for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies \emph{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repositories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline (Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}) also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper.

As far as we know, Physcraper is the only phylogenetic pipeline that links molecular and tree data repositories.
It can also link to various biological databases, such as GBIF, the Paleobiology Database, etc, by leveraging the integrated taxonomy from OpenTree (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), a resource that is contributing to the Encyclopedia of Life main goal of ``exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science'' (Wilson \protect\hyperlink{ref-wilson2003encyclopedia}{2003}).

As such, Physcraper can be used to rapidly (in a matter of hours)
solve challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically (Webb \emph{et al.} \protect\hyperlink{ref-webb2010biodiversity}{2010}),
obtaining trees for ecophylogenetic studies (Helmus \& Ives \protect\hyperlink{ref-helmus2012phylogenetic}{2012}),
sistematizing molecular (and other) databases, i.e., curating taxonomic assignations (San Mauro \& Agorreta \protect\hyperlink{ref-san2010molecular}{2010}),
and generating custom species trees for ecological and evolutionary downstream analyses (Stoltzfus \emph{et al.} \protect\hyperlink{ref-stoltzfus2013phylotastic}{2013}).

Data repositories hold more information than meets the eye.
Besides the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.
Usually, initial ideas about the data are changed by new analyses.
Ideally, new understanding of the data can be registered on databases,
exposing newcomers to the most up to date knowledge about the data.
